Synthesis of Linear Functional Programs

### 2021-07-19 article/article.tex
Fair conjunction ADT destruction + binary restrictions tweaks

\mypara{Core Rules} The core system comprises of proof-search in sequent
calculus rules for linear logic with resource-management and focusing.
%
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove \emph{inessencial} non-determinism from
proof-search by structuring the application of invertible and non-invertible
inference rules. Rules are classified as \emph{left} or \emph{right},
\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
conseguir formular bem esta introdução}
The proof search is separated into two phases:
%
inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later
try to invert, moving them to the linear context ($\Delta$) when we fail to.

@@

The core language is a simply typed linear lambda calculus for which the terms
($M,N$)
and types' ($\tau$) syntax is constructed with the following grammar:
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

>>>>

\mypara{Core Rules} The core system comprises of proof-search in sequent
calculus rules for linear logic with resource-management and focusing.
%
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove inessencial non-determinism from
proof-search by structuring the application of invertible and non-invertible
inference rules. Rules are classified as \emph{left} or \emph{right},
\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
conseguir formular bem esta introdução}
The proof search is separated into two phases:
%
inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later
try to invert, moving them to the linear context ($\Delta$) when we fail to.

@@

The core language is a simply-typed linear lambda calculus with linear
functions, additive and multiplicative pairs, multiplicative unit, additive sums
and the exponential modality (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais?
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.
